---
title       : Stan demonstration
subtitle    : 
author      : Ziyuan Lin
job         : 
framework   : io2012        # {io2012, html5slides, shower, dzslides, ...}
highlighter : highlight.js  # {highlight.js, prettify, highlight}
hitheme     : tomorrow      # 
widgets     : [mathjax, bootstrap]            # {mathjax, quiz, bootstrap}
mode        : selfcontained # {selfcontained, standalone, draft}
knit        : slidify::knit2slides
output: 
  html_document:
    fig_caption: true
  
---



## Why Stan?

> Optimization is bad, expectations are really important, Hamiltonian Monte Carlo is the right thing to do, and Stan is the right way to do the Hamiltonian Monte Carlo. 
> <p style="text-align:right">-- Michael Betancourt, Reykjavik, 2014</p>

## Workflow

1. Write the source code in the modeling language to specify the model ($P(\mathcal{D}\mid\theta)$);
2. Use the compiler or some wrapper to compile and do the inference ($P(\theta\mid\mathcal{D})$).
 + Shell, R, Python, MATLAB, Julia

   We use RStan here.

--- &tabs

## Example 1: linear regression

$$
y_n=\alpha+\beta x_n+\epsilon_n\quad\mbox{where}\quad\epsilon_n\sim\mathcal{N}(0,\sigma^2)
$$

*** class:active id:figure

```{r echo=FALSE, fig.align='center', fig.cap='linear regression'}
library(grid)
library(ggplot2)
n <- 100
x <- 1:n
alpha <- 50
beta <- 0.2
y <- alpha + beta * x + rnorm(n, mean=0, sd=4)
lr_r_data <- data.frame(x=x,y=y)
ggplot(lr_r_data, aes(x=x,y=y)) + geom_point() + theme_bw() + coord_fixed(ratio=2) + stat_smooth(method=lm)
```


*** id:stan-code

```{r eval=FALSE}
data {
  int<lower=0> N;
  vector[N] x;
  vector[N] y;
}
parameters {
  real alpha;
  real beta;
  real<lower=0> sigma_sqrt;
}
model {
  y ~ normal(alpha + beta * x, sigma_sqrt); // no prior = improper prior
}
```

--- &twocol_with_width

## First things to learn

$$
y_n=\alpha+\beta x_n+\epsilon_n\quad\mbox{where}\quad\epsilon_n\sim\mathcal{N}(0,\sigma^2)
$$

*** =left width:50%

```{r eval=FALSE}
data {
  int<lower=0> N;
  // multidimensional: matrix[N,K] x;
  vector[N] x; 
  vector[N] y;
}
parameters {
  real alpha;
  // multidimensional: vector[K] beta;
  real beta; 
  real<lower=0> sigma_sqrt;
}
model {
  y ~ normal(alpha + beta * x, sigma_sqrt); 
}
```

*** =right width:50%

Blocks:
 1. required: `data`, `parameters`, `model`
 2. optional:
   + `transformed data`, <br>`transformed parameters`: <br>"...allow new variables to be declared and then defined through a
sequence of statements."
   + `generated quantities`: <br>for direct generation (e.g., you need to predict some `y_new` given `x_new`)


--- &twocol_with_width

## First things to learn

$$
y_n=\alpha+\beta x_n+\epsilon_n\quad\mbox{where}\quad\epsilon_n\sim\mathcal{N}(0,\sigma^2)
$$

*** =left width:50%

```{r eval=FALSE}
data {
  int<lower=0> N;
  // multidimensional: matrix[N,K] x;
  vector[N] x; 
  vector[N] y;
}
parameters {
  real alpha;
  // multidimensional: vector[K] beta;
  real beta; 
  real<lower=0> sigma_sqrt;
}
model {
  y ~ normal(alpha + beta * x, sigma_sqrt); 
}
```

*** =right width:50%

Data types:
 1. primitive: `int`, `real`
 2. vector/matrix: `vector`, `row_vector`, `matrix`
 3. array: "any type can be made into an array type by declaring array arguments." <br>Examples: <br>`real x[10]; matrix[3,3] m[6,7];`
 4. constraints
  + `upper` and `lower`
  + `unit_vector`, `cov_matrix`, ...


--- &tabs

## Compile and do inference

*** class:active id:R-code

```{r eval=FALSE}
library(rstan)
N <- 100; x <- 1:N; alpha <- 50; beta <- 0.2
y <- alpha + beta * x + rnorm(n, mean=0, sd=4)
lr_stan_data <- list(N=N, x=x, y=y)
# stan source code stored in lr_stan_code
lr_fit <- stan(model_code=lr_stan_code, data=lr_stan_data, iter=1000, chains=4)
```

*** id:stan-result

```{r eval=FALSE}
> lr_fit
Inference for Stan model: lr_code.
4 chains, each with iter=1000; warmup=500; thin=1; 
post-warmup draws per chain=500, total post-warmup draws=2000.

              mean se_mean   sd    2.5%     25%     50%     75%   97.5% n_eff Rhat
alpha        52.16    0.03 0.79   50.56   51.62   52.18   52.72   53.65   679 1.00
beta          0.17    0.00 0.01    0.14    0.16    0.17    0.17    0.19   688 1.00
sigma_sqrt    3.97    0.01 0.29    3.45    3.77    3.96    4.15    4.59   609 1.00
lp__       -185.67    0.06 1.25 -188.86 -186.29 -185.35 -184.77 -184.26   497 1.01

Samples were drawn using NUTS(diag_e) at Sun Mar 22 22:21:52 2015.
For each parameter, n_eff is a crude measure of effective sample size,
and Rhat is the potential scale reduction factor on split chains (at 
convergence, Rhat=1).
```

--- &tabs

## Example 2: Gaussian process

### Younger Americans are more likely to support the statewide legalization of gay marriage

*** class:active id:data

```{r echo=FALSE, fig.align='center'}
gp_stan_data <- list(
  N_ages=80,
  ages=c(18, 19, 20, 21, 22, 23, 24, 25, 26, 27, 28, 29, 30, 31, 32, 33, 34, 35, 36, 37, 38, 39, 40, 41, 42, 43, 44, 45, 46, 47, 48, 49, 50, 51, 52, 53, 54, 55, 56, 57, 58, 59, 60, 61, 62, 63, 64, 65, 66, 67, 68, 69, 70, 71, 72, 73, 74, 75, 76, 77, 78, 79, 80, 81, 82, 83, 84, 85, 86, 87, 88, 89, 90, 91, 92, 93, 94, 95, 96, 97),
  K_male=c(38, 36, 32, 25, 28, 44, 41, 49, 46, 37, 50, 39, 51, 43, 52, 55, 41, 39, 52, 27, 45, 40, 56, 41, 42, 45, 45, 54, 44, 41, 39, 42, 47, 38, 47, 36, 38, 48, 38, 36, 23, 27, 26, 25, 29, 10, 15, 16, 13, 14, 16, 8, 11, 12, 11, 11, 8, 12, 7, 8, 8, 3, 5, 4, 4, 6, 3, 0, 1, 2, 0, 2, 0, 0, 1, 0, 1, 0, 0, 0),
  N_male=c(100, 77, 63, 73, 77, 104, 114, 124, 96, 98, 119, 112, 131, 142, 162, 157, 124, 128, 148, 114, 156, 122, 189, 132, 175, 151, 175, 191, 165, 170, 142, 163, 177, 137, 165, 124, 135, 139, 141, 139, 119, 97, 114, 97, 113, 90, 74, 111, 66, 75, 73, 52, 70, 54, 69, 62, 44, 62, 43, 45, 41, 33, 48, 28, 33, 25, 22, 20, 12, 9, 9, 6, 6, 0, 5, 1, 2, 0, 0, 0),
  K_female=c(62, 40, 31, 57, 55, 60, 64, 82, 62, 69, 71, 50, 75, 59, 55, 85, 70, 59, 55, 77, 66, 51, 79, 53, 70, 66, 60, 81, 80, 80, 66, 65, 93, 74, 60, 70, 70, 66, 58, 65, 38, 48, 44, 33, 27, 26, 25, 42, 17, 20, 21, 10, 29, 18, 22, 26, 16, 24, 20, 15, 6, 8, 16, 4, 11, 4, 6, 8, 5, 1, 2, 1, 2, 0, 0, 1, 1, 0, 0, 1),
  N_female=c(100, 76, 65, 99, 105, 125, 115, 139, 126, 151, 134, 117, 164, 134, 145, 178, 182, 145, 144, 155, 172, 156, 242, 157, 204, 162, 156, 229, 195, 202, 186, 214, 246, 176, 180, 185, 190, 177, 159, 184, 136, 149, 176, 112, 133, 110, 104, 148, 98, 93, 101, 84, 128, 82, 107, 83, 83, 80, 75, 77, 58, 42, 82, 47, 54, 43, 36, 30, 28, 14, 14, 13, 12, 9, 4, 2, 1, 2, 2, 1),
  K=c(100, 76, 63, 82, 83, 104, 105, 131, 108, 106, 121, 89, 126, 102, 107, 140, 111, 98, 107, 104, 111, 91, 135, 94, 112, 111, 105, 135, 124, 121, 105, 107, 140, 112, 107, 106, 108, 114, 96, 101, 61, 75, 70, 58, 56, 36, 40, 58, 30, 34, 37, 18, 40, 30, 33, 37, 24, 36, 27, 23, 14, 11, 21, 8, 15, 10, 9, 8, 6, 3, 2, 3, 2, 0, 1, 1, 2, 0, 0, 1),
  N=c(200, 153, 128, 172, 182, 229, 229, 263, 222, 249, 253, 229, 295, 276, 307, 335, 306, 273, 292, 269, 328, 278, 431, 289, 379, 313, 331, 420, 360, 372, 328, 377, 423, 313, 345, 309, 325, 316, 300, 323, 255, 246, 290, 209, 246, 200, 178, 259, 164, 168, 174, 136, 198, 136, 176, 145, 127, 142, 118, 122, 99, 75, 130, 75, 87, 68, 58, 50, 40, 23, 23, 19, 18, 9, 9, 3, 3, 2, 2, 1)
)

gp_r_data <- data.frame(portion=(gp_stan_data$K_male+gp_stan_data$K_female)/(gp_stan_data$N_male+gp_stan_data$N_female), ages=gp_stan_data$ages)

ggplot(gp_r_data, aes(x=ages,y=portion)) + geom_point() + theme_bw() + ylab("support for gay marriage (K/N)")
```

*** id:gp-result

```{r echo=FALSE, fig.align='center'}
load('code/gp_fit.rdata')
p_post_mean <- sapply(1:gp_stan_data$N_ages, FUN=function(i){
  p_post_chain_mean <- sapply(1:4, function(idx_chain){
    chain <- gp_fit@sim$samples[[idx_chain]]
    list_name <- paste("`p_post[", i, "]`", sep="")
    mean(eval(parse(text=paste("chain$", list_name, sep=""))))
  })
  mean(p_post_chain_mean)
})

p_post_sd <- sapply(1:gp_stan_data$N_ages, FUN=function(i){
  p_post_chain_sd <- sapply(1:4, function(idx_chain){
    chain <- gp_fit@sim$samples[[idx_chain]]
    list_name <- paste("`p_post[", i, "]`", sep="")
    sd(eval(parse(text=paste("chain$", list_name, sep=""))))
  })
  mean(p_post_chain_sd)
})  
  
gp_post_data <- data.frame(ages=gp_stan_data$ages, p_post_mean=p_post_mean, p_post_sd=p_post_sd)
ggplot() + geom_point(data=gp_r_data, aes(x=ages,y=portion)) + geom_line(data=gp_post_data, aes(x=ages,y=p_post_mean)) + geom_ribbon(data=gp_post_data, aes(x=ages, ymin=p_post_mean-p_post_sd, ymax=p_post_mean+p_post_sd), fill="gray", inherit.aes=F, alpha=0.5) + theme_bw() + ylab("support for gay marriage (K/N)")
```
